Case-based Arabic Morphological Analyzer for Arabic Text Mining

### 2010-12-17 trunk/tex/sarf.aclhlt.tex
at 8 pages

\begin{document}
\maketitle
\begin{abstract}
Natural language processing uses morphological 
analysis to abstract and annotate text.
Often times, and in particular in the context
of the Arabic language, annotations resulting from 
an ongoing branch of morphological analysis may not be
appropriate for the case study at hand. 
In this paper we present Sarf, a case based morphological 
analyzer for Arabic text. 
Sarf allows a case-based controller to control and refine
the state of the analysis on the fly. 
It uses non-deterministic finite state Automata
to analyze a stream of text. 
Each alive machine corresponds to a valid analysis. 
Sarf uses concatenative analysis based on recursive affixes 
to better preserve part of speech information.
We automated the analysis of three books of Islamic
narrations (hadith) using Sarf where we
abstracted each book into a vector of complex structures.
Our results show higher accuracy and better efficiency
compared to current analyzers.
\end{quote}
\end{abstract}

@@

Automated analysis of Arabic data sets, including texts, 
publications, records and digital media is essential
with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
Current morphological analyzers~\cite{Sughaiyer:04}
use concatenative analysis
to analyze an Arabic word and
consider its internal structure composed of several
{\em morphemes}. A morpheme can be a {\em stem}, or an {\em affix}.
An affix can be a {\em prefix, suffix, } or an {\em infix}.
The analysis of one word may lead to several possible
morphological solutions.
\vocalize
For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
may have two valid morphological analyses. 
The letter \RL{'a} may be a prefix and the word means 
``I praise him'', or 
%The letter \RL{'a} may also be 
part of the stem \RL{'a.hmad} (a proper noun)
and the word means ``his Ahmad''.

\novocalize
The accuracy of the solutions suffer due to inherent difficulties
of morphological analysis of the Arabic language. 
For example, it is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Another difficulty is that  Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase \transfalse
\RL{il_A\nospace almdrsT} \transtrue
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
without the need of a delimiter such as a space in between. 
This happens because the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words,
referred to as ``run-on'' words~\cite{Buckwalter:04},
occur often, and greatly increase the
difficulty of tokenization.

Current morphological analyzers such as 
Buckwalter~\shortcite{Buckwalter:02},
Beesley~\shortcite{Beesley:01},
SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} exist.
They take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This approach has several problems. 
First, a white space delimited token may have 
more than one word.
Second, the exhaustive enumeration may hurt performance and may
not be necessary or appropriate
in some case studies as noted in~\cite{Maamouri:10}. 
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
also use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of performance.
Xerox rules can be compiled into specialized finite state
machine (FSM) based analyzers as described in~\cite{Beesley:03}.
The efficiency of the resulting analyzer highly depends on the
way the Xerox rules are written. 
This requires deep knowledge and insight from the user
in morphological analysis as well as skills in writing 
context free grammar rules that can be automatically compiled
into efficient non-deterministic finite state Automata (NDFSA).

We hypothesize that many NLP case studies need the 
morphological analyzer to answer queries that do not need 
high accuracy at the low morphological analysis level.
The analysis at the higher levels can often compensate for 
tolerable inaccuracy at low levels. 
For example, if a query concerns proper names and the 
prefix in question in the analysis connects only to a verb, 
we do not need to further analyze the rest of the word 
to provide an answer.

We find evidence to our hypothesis in~\cite{Maamouri:10} where 
a new corpus was added to the Arabic Tree Bank 
(ATB)~\cite{Maamouri:04}. 
The addition required compliance with the DARPA GALE program 
including several NLP tasks such as English translation 
annotations and word-level alignment of Arabic and English. 
The new genre of data challenged the existing data processing and 
annotation guidelines. 
Alternative guidelines came to place where the high level NLP 
task interacted with the morphological analyzer. 
This led to refining SAMA~\cite{Kulick:10}.

@@

\subsection{Sarf}
\label{sec:intro:sarf}

In this paper, we present Sarf\footnote{Sarf (name modified for blind review purposes) 
will be available online as an open source tool.},
 a {\em novel case-based efficient
morphological analyzer} that uses 
non-deterministic finite state Automata 
driven by a case based controller.
%Each alive machine in Sarf represents a valid morphological analysis.
Sarf keeps alive all possible analyses of the text and gives 
opportunity to the controller to intervene at every single input 
letter. 
The decisions of the case-based controller can thus prune false 
positives early in the run. 
Each alive machine in Sarf takes as input one letter at a time 
from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

@@

We validate our hypothesis using a case study from the Islamic 
literature. 
A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
Arabic with its transliteration and translation. 
The authentication of a \noArRL{.hady_t} depends on 
the credibility of each of the narrators as reported in 
separate biography books. 
In this paper we consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, then segmenting each 
narration into its content or \RL{matn} and its \noArRL{sanad}.
We also partition the sanad into the 
separate narrators so that we can later look each one of them 
up in the biography books. 
For example, the boxes in 
Figure~\ref{f:exhadith} are proper names and once
connected together they form complex names of narrators. 
For example, narrator $n_1$ has the first name 
\noTrRL{qtybT} and his father name is
\noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.

The hadith case study is interesting to Sarf since the controller
considers most of the analyses where we have a concentration
of proper names in the text. It ignores the text where
proper names occur rarely.% as long 
%as the text is valid.
We are also interested in discovering a subset of words 
that relate persons to each other such as \RL{bin} or ``the son of''
or words that mean narrate such as \RL{qaal} or ``said''. 
With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of complex structures 
that are three level deep in hierarchy.
%We also showcase that the results of the accurate high level 
%abstraction
%results of the case-based analysis was not affected by a major
%refinement to the analyzer that increased low level analysis accuracy.

%\subsection{Contributions}

We make the following key contributions. 
\begin{enumerate}
\item {\bf Case-based analysis:}  We designed and implemented
a case-based morphological analyzer where a case-based controller
machine controls and guides the analysis. 
%\item {\bf Exhaustive analysis:} Since we have a case-based 
%controller that can decide on the fly whether an analysis is 
%accepted or not, we can afford to keep all valid analyses modulo
%those pruned by the case controller. 
%We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02} and define the affixes
recursively. 
This allows Sarf to perform better and 
maintain compositional part of speech tags.
\item {\bf Hadith case study:}
We evaluated Sarf using the hadith case study. We were able to
efficiently extract the sequence of narrators into a complex
data structure with three levels of hierarchy with high accuracy. 
\end{enumerate}

@@

The rest of this paper is structured as follows. In Section~\ref{sec:related}
we compare Sarf to related work. Then in Section~\ref{sec:sarf}
we discuss Sarf and illustrate the way it works with a running example.
We also explain how non-deterministic finite state Automata
work. In Section~\ref{sec:islamic} we present the hadith case
study. We present our results in Section~\ref{sec:results}
and conclude with future work in Section~\ref{sec:future}.

@@

The survey in~\cite{Sughaiyer:04} discusses and compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific applications in the morphological 
analyzer itself or use a specific set of part of speech tags
as their reference.
Sarf differs in that it is a general and rich morphological 
analyzer that reports all possible solutions. 
It is case-based in the sense that a controller that drives
the morphological analyzer prunes these decisions.

Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
It differs from Buckwalter in that it defines a shorter list of affixes
and a longer list of affix compatibility rules to allow recursive 
affixes so that we can have prefix-prefix an suffix-suffix 
concatenations.
This allows Sarf to better maintain and propagate 
the part of speech information
tags associated with each prefix and suffix.
This information is highly 
redundant with Buckwalter.% and that may lead to consistency issues. 
Sarf keeps all the possible analyses alive where each analysis
corresponds to a set of decisions and positions in the text. 
Buckwalter produces a set of solutions for each token 
and the several analysis threads of the text need to be 
generated as a product of the token solutions.

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with an FSA that is constructed from 
the controller and 
the prefix, stem, and suffix deterministic transducers. 
Sarf differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller. 
With Sarf it is easy to control the FSA with a custom user case 
controller. Up to our understanding, Beesley  
requires recompiling the Xerox rules.

Sarf is similar to The MADA+TOKAN toolkit in that it thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
case study to interfere in judging the analyses much earlier and
at a finer granularity level. 
In Sarf, there is no need for a separate tokenizer such as
TOKAN since each solution keeps a stack of positions
that partition text into morphemes.
%The case-based controller can build the trace 
%when needed.

@@

Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. 
%We find supporting evidence in this decision to our hypothesis
%about the utility of a case-based analyzer. 
The algorithmic changes in SAMA were
done manually and worked in integration with the ATB format. 
Sarf is a case-based controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with any case study. 
In Sarf, the analyzer remains general, and the 
case-based controller specializes the analysis.

The AMIRA~\cite{Diab:07} analyzer uses 
a language independent SVM based analysis. 
While the SVM analysis learns features from the text passed to the case
study, it does not make use of the type of the query the 
case study targets in the text.

@@

Sarf encodes the prefix, suffix and stem lexicons into
three linear transducers and uses a non-deterministic construction
of the three to compute all valid morphological analyses.
We first present a running example of Sarf.
The running example explains how Sarf works,
implements recursive affixes, and
handles ``run-on'' words.
%Then we discuss how Sarf builds and composes the different FSAs.

@@

The diagram in Figure~\ref{f:example}
illustrates the operation of $\Phi$ the Sarf FSA 
when parsing the
string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
    \RL{n}%
        \RL{w}
    \RL{b}
    \RL{`}
    \RL{a} 
    \RL{l}
    \RL{l}
    \RL{a} %
        \RL{a}
    \RL{h}
    \RL{b}
    \RL{`}
    \RL{l}
    \RL{y}
    \RL{s}
    \RL{w}
    in separate form to ease following the example
}

The box and circle denote accept and regular
states respectively.
The edges are transitions with labels corresponding to
the input letters.
The reject states are omitted for clarity and are denoted
by the absence of corresponding transitions.

@@

We start with $\Phi$ at the root square state in 
${\cal P}$ which is an accept state. 
The $\epsilon$-edge connects the ${\cal P}$
to ${\cal S}$ and transitions from any accept state
in ${\cal P}$ to the root state in ${\cal S}$.
The $\epsilon$-edge that connects the ${\cal S}$ to ${\cal X}$
follows the same behavior.

When there are two valid transitions such as \RL{w} 
and $\epsilon$ in the start case, 
$\Phi$ spawns an exact copy of itself $\Psi$. 
$\Phi$ makes one transition, and $\Psi$ makes the other. 
Each of $\Phi$ and $\Psi$ now represent a valid analysis so far. 
An FSA ($\Phi$ or $\Psi$) dies when it reaches a reject state.
In our example, if there were no stems that start 
with the letter \RL{w}, $\Psi$ will die. 
In reality, $\Psi$ will die when the input 
is at \RL{wsyl`}.

Note that the affix transducers allow recursive 
affixes. 
For example, \RL{w} will result in an accept state
that transitions to the ${\cal S}$.
When \RL{s} follows, we move to another accept state in 
${\cal P}$ corresponding to the prefix \RL{ws}. 
The same applies to ${\cal X}$.

@@

Before moving with the letter \RL{b} to the accept state,
$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
with the prefix \RL{wsy}. 
It does that by associating a category value for each accept state. 
Sarf considers the value of the category as part of the state.
Thus each state shown in Figure~\ref{f:example} actually represents
more than one state. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

Since $\Phi$ is now in an accept state in ${\cal S}$, it will
spawn $\Xi$ that transitions into the root of ${\cal X}$. 
When \RL{h} is consumed, $\Phi$ will die since 
there is no edge from the current state labeled with \RL{h}. 
Now $\Xi$ represents a valid analysis and can proceed.

@@

The case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an FSA that 
is not interesting. 
It may also decide to correct the analysis of one FSA
based on the decision of the others. 
The controller uses the output of the transducers as input
to make its decision. 
It does not produce external output and therefore 
the three transducers with the 
controller compose an FSA.

@@

We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\shortcite{Buckwalter:02} keeps separate lists 
of affix and stem morphemes and assigns categories to
morphemes. 
%The relation between morphemes and categories is one 
%to many. 
A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that morphemes
with category $c_1$ can be concatenated with morphemes
of category $c_2$. 
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
The $\alpha\beta\gamma$ partition of $s$ is a 
valid morphological analysis if and only if 
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$, where
$c()$ returns the category of the morpheme.

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The part of speech (POS) tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping all valid morphemes in a list
is redundant and may lead to consistency issues when
users try to add their custom morphemes.
We introduce $L_{pp}$ and
$L_{xx}$ as prefix-prefix and suffix-suffix 
compatibility rules respectively.
This allows us to keep a shorter list of affixes. 
We also introduce a {\em resulting category}
for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
The resulting category in all practical cases
is that of the second morpheme in the pair. 
Atomic affixes are standalone affixes
that are not the concatenation of other affixes. 
Formally, we define an affix $f: a | a f$ to be either
an atomic affix $a$ or an atomic affix concatenated
with an affix. 
%Sarf only keeps atomic affixes and their categories.

@@

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and constructed the recursive affixes
into directed acyclic graphs (DAG) similar to 
those in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
This transducer is deterministic and thus 
its traversal is linear.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings %of the word in question
and looks them up in the affix tables.

@@

\subsection{Comparison with Buckwalter and ElixirFM}

>>>>

\begin{document}
\maketitle
\begin{abstract}
Natural language processing uses morphological 
analysis to abstract and annotate text.
Often times, 
%      and in particular in the context of the Arabic language, 
annotations resulting from 
a branch of morphological analysis may not be
appropriate for the case study at hand. 
In this paper we present Sarf, a case based morphological 
analyzer for Arabic. 
Sarf allows a case-based controller to control and refine
the state of the analysis on the fly. 
It uses non-deterministic finite state Automata
to analyze a stream of text. 
Each alive machine corresponds to a valid analysis. 
Sarf uses concatenative analysis based on recursive affixes 
to better preserve part of speech information.
We automated the analysis of three books of Islamic
narrations (hadith) using Sarf where we
abstracted each book into a vector of complex structures.
Our results show higher accuracy and better efficiency
compared to current analyzers.
\end{quote}
\end{abstract}

@@

%Automated analysis of Arabic data sets, including texts, 
%publications, records and digital media is essential
%with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
Current morphological analyzers~\cite{Sughaiyer:04}
use concatenative analysis
consider the internal structure 
of an Arabic word and
compose it into several {\em morphemes}. 
A morpheme can be a {\em stem}, or an {\em affix}.
An affix can be a {\em prefix, suffix, } or an {\em infix}.
The analysis of one word may lead to several possible
morphological solutions.
\vocalize
For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
may have two valid morphological analyses. 
The letter \RL{'a} may be a prefix and the word means 
``I praise him'', or 
%The letter \RL{'a} may also be 
part of the stem \RL{'a.hmad} (a proper noun)
and the word means ``his Ahmad''.

\novocalize
The accuracy of the solutions suffer due to inherent difficulties
of morphological analysis of the Arabic language. 
For example, it is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase \transfalse
\RL{il_A\nospace almdrsT} \transtrue
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
without the need of a space in between. 
The reason is the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words,
referred to as ``run-on'' words~\cite{Buckwalter:04},
occur often, and greatly increase the
difficulty of tokenization.

Current morphological analyzers such as 
Buckwalter~\shortcite{Buckwalter:02},
Beesley~\shortcite{Beesley:01},
SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} 
take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This approach has several problems. 
First, a white space delimited token may have 
more than one word.
Second, the exhaustive enumeration may hurt performance and may
not be necessary or appropriate
in some case studies as noted in~\cite{Maamouri:10}. 
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of performance.
Xerox rules can be compiled into specialized finite state
machine (FSM) based analyzers as described in~\cite{Beesley:03}.
The efficiency of the resulting analyzer depends on the
way the Xerox rules are written. 
This requires deep knowledge and insight from the user
in compilation techniques, context free grammars, 
and morphological analysis.
%context free grammar rules that can be automatically compiled
%into efficient non-deterministic finite state Automata (NDFSA).

We hypothesize that many NLP case studies need the 
morphological analyzer to answer queries that do not need 
high accuracy at the low morphological analysis level.
The analysis at the higher levels can often compensate for 
tolerable inaccuracy at low levels. 
For example, if a query concerns proper names and the 
analyzer is considering a 
prefix that connects only to a verb,
the analyzer can answer without considering the 
rest of the word.

We find evidence to our hypothesis in~\cite{Maamouri:10}.
The addition of a new corpus to the Arabic Tree Bank 
(ATB)~\cite{Maamouri:04}
required compliance with the DARPA GALE program 
including several NLP tasks such as English translation 
annotations and word-level alignment of Arabic and English. 
The new data challenged the existing processing and 
annotation guidelines. 
This lef to refining 
SAMA~\cite{Kulick:10} with 
alternative guidelines where the NLP 
task interacted with the morphological analyzer.

@@

%\subsection{Sarf}
%\label{sec:intro:sarf}

{\bf Sarf.~~~}
In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
will be available online as an open source tool.},
 a {\em novel case-based efficient
morphological analyzer} that uses 
non-deterministic finite state Automata 
driven by a case-based controller.
Each alive machine in Sarf represents a valid morphological analysis
and takes as input one letter at a time.
The case-based controller prunes false positives
early in the run by making a decision at every input letter
to control the alive machines.
%Each alive machine in Sarf takes as input one letter at a time 
%from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

@@

We validate our hypothesis using a case study from the Islamic 
literature. 
A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
Arabic with its transliteration and translation. 
The authenticity of a \noArRL{.hady_t} depends on 
the credibility of the narrators as reported in 
separate biography books. 
In this paper, we consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, each 
narration into its \RL{matn} (content) and its \noArRL{sanad},
and each sanad into the 
separate narrators. Now, we can look each narrator 
up in the biography books. 
For example, 
Figure~\ref{f:exhadith} shows proper names in boxes connected
to form complex names of narrators. 
For example, 
\noTrRL{qtybT} is the first name 
of narrator $n_1$, and 
\noTrRL{s`yd} is the name of his father as 
the word \noTrRL{bn} (son of) indicates.

The hadith case study is interesting to Sarf since the controller
considers %most of the analyses where we have a concentration
proper names and ignores text where
proper names occur rarely.% as long %as the text is valid.
We are also interested in discovering a words 
that relate names to each other such as \RL{bin} or ``the son of''
and words that mean narrate such as \RL{qaal} or ``said''.

With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of complex structures 
that are three level deep in hierarchy.
%We also showcase that the results of the accurate high level 
%abstraction
%results of the case-based analysis was not affected by a major
%refinement to the analyzer that increased low level analysis accuracy. 
We make the following key contributions. 
\begin{enumerate}
\item {\bf Case-based analysis:}  We designed and implemented
a case-based morphological analyzer where a controller
guides the analysis. 
%\item {\bf Exhaustive analysis:} Since we have a case-based 
%controller that can decide on the fly whether an analysis is 
%accepted or not, we can afford to keep all valid analyses modulo
%those pruned by the case controller. 
%We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02} and define affixes
recursively. 
This allows Sarf to perform better and 
maintain compositional part of speech (POS) tags.
\item {\bf Hadith case study:}
%We evaluated Sarf using the hadith case study. 
We were able to efficiently extract a complex three 
level hierarchical 
structure containing the chains of narrators 
with high accuracy. 
\end{enumerate}

@@

The rest of this paper is structured as follows. In Section~\ref{sec:related}
we compare Sarf to related work. In Section~\ref{sec:sarf}
we discuss Sarf and illustrate it with a running example.
%We also explain how non-deterministic finite state Automata
%work. 
In Section~\ref{sec:islamic} we present the hadith case
study. We present our results in Section~\ref{sec:results}
and conclude with future work in Section~\ref{sec:future}.

@@

The survey in~\cite{Sughaiyer:04} compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific applications in the 
analyzer itself or use a specific set of POS tags
as their reference.
Sarf differs in that it is a general morphological 
analyzer that reports all possible solutions. 
It is case-based in the sense that a controller 
prunes these solutions.

Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
It differs from Buckwalter in that it defines a shorter list of affixes
and a longer list of affix compatibility rules to allow recursive 
affixes so that we can have prefix-prefix an suffix-suffix 
concatenations.
This allows Sarf to better maintain and propagate 
the POS tags associated with affixes. 
%This information is highly 
%redundant with Buckwalter.% and that may lead to consistency issues. 
Sarf keeps all the possible analyses alive where each analysis
corresponds to a set of decisions and positions in the text. 
Buckwalter produces a set of solutions for each token 
and the several analysis threads of the text need to be 
generated as a product of the token solutions.

Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
finite state Automata (FSA). 
It reports that the number of Automata generated by a compiler
for Xerox rules can not be controlled and also reports a difficulty 
to compose the FSAs into a single framework. 
Sarf constructs an NDFSA from
the controller and the prefix, stem, and suffix deterministic 
transducers to solve the problem.
Sarf differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller to modify their behavior. 
%With Sarf it is easy to control the FSA with a case based
%controller. 
Beesley requires recompilation.% of the Xerox rules.

Similar to The MADA+TOKAN toolkit, Sarf thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
case study to interfere in judging the analyses much earlier and
at a finer granularity level. 
In Sarf, there is no need for a separate tokenizer such as
TOKAN since each solution keeps a stack of positions
that partition text into morphemes.
%The case-based controller can build the trace 
%when needed.

@@

Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. 
%We find supporting evidence in this decision to our hypothesis
%about the utility of a case-based analyzer. 
The algorithmic changes in SAMA were
done manually and worked in integration with the ATB format. 
Sarf is a case-based controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with any case study. 
%In Sarf, the analyzer remains general, and the 
%case-based controller specializes the analysis.

The AMIRA~\cite{Diab:07} analyzer uses 
a language independent SVM based analysis. 
The SVM analysis learns features from the input text 
and does not make use of any information 
in the target query.

@@

Sarf encodes the prefix, suffix and stem lexicons into
three deterministic transducers and 
uses a non-deterministic construction
of the three to compute all valid morphological analyses.
We first present a running example of Sarf.
The running example explains how Sarf works,
implements recursive affixes, and
handles ``run-on'' words.
%Then we discuss how Sarf builds and composes the different FSAs.

@@

The diagram in Figure~\ref{f:example}
illustrates the operation of $\Phi$, the Sarf FSA,
when parsing the
string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
\RL{n}%
\RL{w}
\RL{b}
\RL{`}
\RL{a} 
\RL{l}
\RL{l}
\RL{a} %
\RL{a}
\RL{h}
\RL{b}
\RL{`}
\RL{l}
\RL{y}
\RL{s}
\RL{w}
in separate form to ease following the example
}

Boxes and circles denote accept and regular
states respectively.
The edges are transitions with labels corresponding to
the input letters.
The reject states are omitted for clarity and are denoted
by the absence of corresponding transitions.

@@

We start with $\Phi$ at the root square state in 
${\cal P}$ which is an accept state. 
The $\epsilon$-edge connects ${\cal P}$
to ${\cal S}$ and transitions from any accept state
in ${\cal P}$ to the root state in ${\cal S}$.
The $\epsilon$-edge that connects ${\cal S}$ to ${\cal X}$
follows the same behavior.

When there are two valid transitions such as \RL{w} 
and $\epsilon$ in the start case, 
$\Phi$ spawns an exact copy of itself $\Psi$. 
$\Phi$ makes one transition, and $\Psi$ makes the other. 
Each of $\Phi$ and $\Psi$ represent a valid analysis so far. 
An FSA ($\Phi$ or $\Psi$) dies when it reaches a reject state.
In our example, if there were no stems that start 
with the letter \RL{w}, $\Psi$ will die. 
In reality, $\Psi$ will die when the input 
is at \RL{wsyl`}.

Note that the affix transducers allow recursive 
affixes. 
For example, \RL{w} will result in an accept state
that transitions to ${\cal S}$.
When \RL{s} follows, we move to another accept state in 
${\cal P}$ corresponding to the prefix \RL{ws}. 
The same applies to ${\cal X}$.

@@

Before moving with the letter \RL{b} to the accept state,
$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
with the prefix \RL{wsy}. 
Sarf keeps a compatibility category values as part
of the accept states. 
%Sarf considers the value of the category as part of the state.
Thus each accecpt state in Figure~\ref{f:example} represents
more than one concrete state. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

Since $\Phi$ is now in an accept state in ${\cal S}$, it will
spawn $\Xi$. $\Xi$ transitions to the root of ${\cal X}$. 
$\Phi$ die when it consumes \RL{h} since
no edge from the current state is labeled with \RL{h}. 
$\Xi$ represents a valid analysis and can proceed.

@@

The case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an FSA that 
is not interesting. 
It may also decide to correct one FSA
based on the decision of the others. 
The controller uses the output of the transducers as input
to make its decision. 
It does not produce external output and therefore 
the three transducers with the 
controller compose an FSA.

@@

We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\shortcite{Buckwalter:02} keeps separate lists 
of affix and stem morphemes and assigns categories to
morphemes. 
%The relation between morphemes and categories is one 
%to many. 
A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that $c_1$ morphemes
can be concatenated with $c_2$ morphemes.
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix;
$\alpha\beta\gamma$ is a 
valid morphological analysis iff
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$, where
$c()$ returns the category of the morpheme.

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The POS tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping all valid morphemes in a list
is redundant and may lead to consistency issues when
users try to add their custom morphemes.
We introduce $L_{pp}$ and
$L_{xx}$ as prefix-prefix and suffix-suffix 
compatibility rules respectively.
%This allows us to keep a shorter list of affixes. 
We also introduce a {\em resulting category}
for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
%The resulting category in all practical cases
%is that of the second morpheme in the pair.

Atomic affixes are standalone affixes
that are not the concatenation of other affixes. 
Formally, we define an affix $f: a | a f$ to be either
an atomic affix $a$ or an atomic affix concatenated
with an affix. 
%Sarf only keeps atomic affixes and their categories.

@@

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and constructed the recursive affixes
into directed acyclic graphs (DAG) similar to 
those in Figures~\ref{f:example}(a) and~\ref{f:example}(c).
The transducers are deterministic and thus 
their traversal is linear.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings %of the word in question
and looks them up in the affix tables.

@@

\subsection{Buckwalter and ElixirFM}
